---
title: "R Notebook for Imaging Stats MIC Course"
output:
  rmdformats::downcute: default
---

# Load required libraries

```{r setup, message=FALSE, warning=FALSE}
# Fast data processing with data.table
library(data.table)

# Modelling of the dose response
library(drc)

# Print pretty tables
library(kableExtra)

# Create beautiful plots
library(ggplot2)

# Extras for plotting with ggplot2
library(scales)
library(ggthemes)

# Interactive plots
#library(plotly)
```


# Load data

Column names include spaces. Use the `check.names` parameter to change them to proper names.

```{r}
dt = data.table::fread("BBBC013_analyzed_full.csv", check.names = T)

dt %>%
  head %>%
  kbl() %>%
  kable_paper(full_width = F)
```

# Data preparation

Add a grouping column to differentiate between wells A-D and E-H, and set the order of factors in the `drug.type` column.

```{r}
# Add a grouping column
dt[,
   group := ifelse(well.row %in% LETTERS[1:4],
                   "GR1",
                   "GR2")]

# Set the order of factors
dt[,
   drug.type := factor(drug.type, 
                       levels = c("Empty",
                                  "Negative Control",
                                  "Wortmannin",
                                  "LY294002",
                                  "Positive Control"))]
```


# Data overview

```{r, fig.width=8, fig.height=6}
dodge <- position_dodge(width = 0.4)

p1 = ggplot2::ggplot(data = dt,
                     aes(x = as.factor(amount),
                         y = intensity)) +
  ggplot2::geom_violin(aes(fill = drug.type), 
                       position = dodge) +
  ggplot2::geom_boxplot(aes(color = drug.type), 
                        fill = "white",
                        notch = T, 
                        outlier.shape = NA, 
                        position = dodge, 
                        width = 0.1) +
  #ggplot2::scale_x_continuous(trans = scales::pseudo_log_trans(0.1, 10)) +
  ggthemes::scale_fill_tableau(name = "Drug type",
                               palette = "Tableau 10") +
  ggthemes::scale_color_tableau(name = "Drug type",
                                palette = "Tableau 10") +
  ggplot2::facet_wrap(~group, ncol = 1, scales = "free_x") +
  ggplot2::xlab("Concentration") +
  ggplot2::ylab("Intensity") +
  ggplot2::theme_bw()

# Interactive plot
#ggplotly(p1)
p1
```

## Sample size

The number of measurements per group.

```{r}
dt[,
   .N,
   by = .(group,
          drug.type)] %>%
  kbl() %>%
  kable_paper(full_width = F)
```

## Area vs intensity

```{r, fig.width=8, fig.height=6}
ggplot2::ggplot(dt,
                aes(x = area,
                    y = intensity)) +
  ggplot2::geom_point(alpha = 0.05) +
  ggplot2::theme_classic()
```


# Normalisation

Normalised Percentage Activation: $$NPA = \frac{meas - CTRL_-}{CTRL_+ - CTRL_-}$$

Where $CTRL_-$ and $CTRL_+$ are mean negative and positive controls, respectively.

```{r}
#  Add a column with the mean intensity of the negative control for each group
dt[,
   mn.negctrl := mean(.SD[drug.type == "Negative Control",
                          intensity]),
   by = group]


# Add a column with the mean intensity of the positive control for each group
dt[,
   mn.posctrl := mean(.SD[drug.type == "Positive Control", 
                          intensity]),
   by = group]

# Calculate NPA
dt[,
   npa := (intensity - mn.negctrl) / (mn.posctrl - mn.negctrl)]

# Clean; remove temporary columns
dt[,
   c("mn.negctrl", "mn.posctrl") := NULL]
```

Plot of the normalised intensity.

```{r, fig.width=8, fig.height=6}
ggplot2::ggplot(data = dt,
                aes(x = as.factor(amount),
                    y = npa)) +
  ggplot2::geom_violin(aes(fill = drug.type), 
                       position = dodge) +
  ggplot2::geom_boxplot(aes(color = drug.type), 
                        fill = "white",
                        notch = T, 
                        outlier.shape = NA, 
                        position = dodge, 
                        width = 0.1) +
  #ggplot2::scale_x_continuous(trans = scales::pseudo_log_trans(0.1, 10)) +
  ggthemes::scale_fill_tableau(name = "Drug type",
                               palette = "Tableau 10") +
  ggthemes::scale_color_tableau(name = "Drug type",
                                palette = "Tableau 10") +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::facet_wrap(~group, ncol = 1, scales = "free_x") +
  ggplot2::xlab("Concentration") +
  ggplot2::ylab("NPA") +
  ggplot2::theme_bw()
```

# Fit a sigmoid response

Calculate EC50, i.e. concentration that induces a response halfway between the baseline and the maximum.

[Dose-Response Analysis Using R](https://doi.org/10.1371/journal.pone.0146021), PLoS ONE 2015.

Using the [drc package](https://cran.r-project.org/web/packages/drc/).

[Dose Response Modelling in R](http://www.darrenkoppel.com/2020/09/04/dose-response-modelling-and-model-selection-in-r/)



# 3-parameter Hill function

The three-parameter log-logistic dose response function with lower limit 0 is: $$f(x) = \frac{d}{1+\exp(b(\log(x)-\log(e)))}$$

Where:

- $b$ is the steepness of the curve (~ nthe Hill coefficient),
- $d$ is the maximum value,
- $e$ is the dose of the sigmoid midpoint,
- $x$ is the dose.

```{r}
b = -2
d = 2
e = 2

mySigF = function(x) { d / ( 1 + exp(b * ( log(x) - log(e) )) ) }

xarg = 10^seq(-3,3,0.1)

{
  plot(xarg, mySigF(xarg), type = "l", log = "x")
  abline(v = e, lty = 2)
}
```

# Fitting

The default non-robust least squares estimation ("mean").


```{r}
drfit = drc::drm(formula = npa ~ amount,
                 curveid = group,
                 data = dt[drug.type %in% c("Negative Control", "LY294002", "Wortmannin")],
                 fct = LL.3())

summary(drfit)
```

## Fitted model

```{r}
plot(drfit, 
     broken = T, 
     col = T,
     conName = "CTRL-",
     ylim = c(0,2), 
     xlab = "Concentration",
     ylab = "NPA")
```

## Fitted ED50

The estimation of the effective dose (EC50 / ED50 / IC50), i.e. the dose that induces the half-maximum response.

```{r}
drc::ED(drfit, 50, interval = "delta")
```

## Relative potency

Relative potency calculated as the ratio of ED50 estimates for both groups.

```{r}
drc::EDcomp(drfit, c(50, 50), interval = "delta")
```

## Slope comparison 

Comparison of slopes by means of a z-test for estimates from both groups.

```{r}
drc::compParm(drfit, "b", "-")
```



